\sys: Light-touch Data-Parallel Shell Processing

### 2020-05-20 pldi/draft.tex
Stash-half pass on 5

Given a dataflow graph representing a shell program, the \sys
optimizer outputs an optimized dataflow graph in which data
parallelism opportunities have been exploited.  More concretely, the
optimizer starts by splitting input files in several chunks, given the
desirable output dataflow graph width. It then starts from the source
nodes of the graph, iteratively performing the graph transformations
that were described in \Cref{ir:transformations}.
%% If the incoming edges of the node are less than the desirable graph
%% width, it performs edge splits, and then it performs maximal node
%% splits.
The final graph can then be implemented by spawning a process for each
node, and redirecting the inputs and outputs according to the graph
edges.

@@

\noindent
\TODO{Problem: Cat (and other commands that have multiple inputs) is
  not eager. Illustration: Show a parallel pipeline having a cat that
  gathers input somewhere (and a sort merge maybe). The way pipes
  work, this graph will essentially evaluate almost completely
  sequentially. Maybe go through the steps that cat asks for the first
  line of its first input until ithe first input is done and then goes
  to the second. This leads to processor under-utilization. Solution:
  We add eager nodes that eagerly consume all their input (even if
  their output does not ask for it). Eager nodes are themselves part
  of the dataflow graph (and model) and essentially perform an
  identity function. This makes all processors as utilized as possible
  and for CPU heavy tasks it leads to speedup. For IO heavy (and CPU
  light) tasks it has some overhead. We evaluate eager in section X
  (We should also say something about a CPU light task). A good thing
  is that they work for all possible commands with almost no overhead,
  without having to make each specific command eager.}

\begin{figure}[t]
\centering
% https://docs.google.com/drawings/d/13Zx3a9bWi1RvpOgNCSiXirp9UOte-XuaJQBi_lmmPMc/edit
\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_eager.pdf}}
\caption{
  \textbf{Eager primitive.}
  Describe three problems:
    (i) FIFOs alone cannot solve it as they’re blocking;
    (ii) intermediary files alone lose data, as the consumer reaches EOF before the producer; and
    (iii) writing to files + wait does not allow trask-based parallelism.
  Eager is an elegant solution that solves the problem while remaining within
  the model. Describe.
}
\vspace{-15pt}
\label{fig:eager}
\end{figure}

\heading{Readjusting Graph Width}

>>>>

After translating the parallelizable regions of the input script to
dataflow graphs and optimizing these graphs using the transformations
described in \Cref{ir:transformations}, \sys implements the graph by
translating it back to a shell script. In this section we describe how
\sys achieves that, and several technical challenges that it addresses.

%% Given a dataflow graph representing a shell program, the \sys
%% optimizer outputs an optimized dataflow graph in which data
%% parallelism opportunities have been exploited.  More concretely, the
%% optimizer starts by splitting input files in several chunks, given the
%% desirable output dataflow graph width. It then starts from the source
%% nodes of the graph, iteratively performing the graph transformations
%% that were described in \Cref{ir:transformations}.
%% %% If the incoming edges of the node are less than the desirable graph
%% %% width, it performs edge splits, and then it performs maximal node
%% %% splits.
%% The final graph can then be implemented by spawning a process for each
%% node, and redirecting the inputs and outputs according to the graph
%% edges. In this section, we describe several technical challenges that
%% we addressed.

@@

The shell is uniquely lazy, in the sense that most commands and pipes
consume their inputs only when they are ready to process more. This
allows the underlying system to be able to support concurrently
executing many more commands than the processing threads of the
underlying system, by scheduling them concurrently, allowing them to
have progress one at a time. This design decision however, severely
restricts parallelism since nodes are often blocked when their
consumers are processing their input. Consider as an example the
following script:

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  mkfifo t1 t2
  grep "foo" > t1 & grep "foo" > t2 &
  cat t1 t2
\end{lstlisting}

The \ttt{cat} command will only try to consume input from \ttt{t2}
when it is done reading from \ttt{t1}. This means that the second
\ttt{grep} will block without processing any input until the first
\ttt{grep} is done processing. Instead of addressing this issue for
specific commands, \sys inserts and instants several eager \ttt{relay}
nodes in critical points of the pipeline as seen in
\Cref{fig:eager}. These nodes force their producers to produce output
and not block, while also preserving the existing task-based
parallelism---that would not be preserved say with the \todo{third}
solution in \Cref{fig:eager}.

%% \TODO{Problem: Cat (and other commands that have multiple inputs) is
%%   not eager. Illustration: Show a parallel pipeline having a cat that
%%   gathers input somewhere (and a sort merge maybe). The way pipes
%%   work, this graph will essentially evaluate almost completely
%%   sequentially. Maybe go through the steps that cat asks for the first
%%   line of its first input until ithe first input is done and then goes
%%   to the second. This leads to processor under-utilization. Solution:
%%   We add eager nodes that eagerly consume all their input (even if
%%   their output does not ask for it). Eager nodes are themselves part
%%   of the dataflow graph (and model) and essentially perform an
%%   identity function. This makes all processors as utilized as possible
%%   and for CPU heavy tasks it leads to speedup. For IO heavy (and CPU
%%   light) tasks it has some overhead. We evaluate eager in section X
%%   (We should also say something about a CPU light task). A good thing
%%   is that they work for all possible commands with almost no overhead,
%%   without having to make each specific command eager.}

\begin{figure}[t]
\centering
% https://docs.google.com/drawings/d/13Zx3a9bWi1RvpOgNCSiXirp9UOte-XuaJQBi_lmmPMc/edit
\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_eager.pdf}}
\caption{
  \textbf{Eager primitive.}
  Describe three problems:
    (i) FIFOs alone cannot solve it as they’re blocking;
    (ii) intermediary files alone lose data, as the consumer reaches EOF before the producer; and
    (iii) writing to files + wait does not allow trask-based parallelism.
  Eager is an elegant solution that solves the problem while remaining within
  the model. Describe. \TODO{Combine this figure with the above code maybe.}
}
\vspace{-15pt}
\label{fig:eager}
\end{figure}

\heading{Split node implementation}
